\large{Vipassana for Hackers} \\ \Huge{Paper Four: Mastering Spacetime} \\ \large\textit{Version 0.1}

### 2019-12-26 paper-four/paper-four.tex
mastery over spacetime

- mind / body === brain / nervous system

- vipassana the largest superset

- anapana covers time, vipassana covers space

Anapana eliminates space. Continuity is the source of constantly-narrowing attention on space.

It says “focus here, on this tiny area… make the area as tiny as possible, with no
limit or exception. Now all you have to do is follow your breath in the tiniest
slices possible, making time stretch out longer and longer with every sub-slice.”

Anapana is mastery over time.

Vipassana is harder. It stops ignoring space and actually adds it to the equation. You move your attention throughout your entire body (which is the only thing you have direct access to: sensation) with the narrow focus you trained yourself to achieve in the first 3 days of the course. As your apparent physical existence begins to break apart, you’ll open up your attention to the whole body, all at once (or as much as possible at once) but even after 9 Vipassana courses, I rarely ever do anything approximating that because I’m quite a beginner. 🙂

Anyway.

Vipassana is mastery over time + space.

...once you have that, there’s nothing left.

- Day 8: Meditate all the time.

>>>>

The completeness of a meditation can only be understood based on three criteria:

\begin{itemize}
  \item All time: It must cover all of time, for the smallest unit of time
  \item All space: It must cover all of space, for both the smallest unit of space and the largest
  \item Continuity: It must be practiced continuously, in every moment of life
\end{itemize}

In this way, we have a meditation which covers all of time and all of space, for all
available time. No ``largest unit of time'' is requested as time, directly perceived,
can only be of the present moment --- and the present moment is infinitesimally small
by its very nature. In place of the ``largest unit of time'', all moments of
perceived time are to be captured: a total continuity.

In this section we will examine how, even within the 10-day Vipassana course, the
teaching of Vipassana actually covers the largest possible superset of all other
meditation techniques.

\subsubsection{Anapana}

Anapana, as taught in the 10-day course, Satipatthana Sutta course, and 20-day
course, demands that the meditator make an attempt at \textit{eliminating space}: the
attention placed below the nostrils is to become narrower and narrower as awareness
and attention become more acute.

Item 3, continuity, already makes an appearance in the practice of Anapana. In
addition to narrowing the area of awarenss, a secondary goal of Anapana Meditation is
to fix the attention to this narrow area for as long a period as possible. Initially,
this continuity is understood in terms of mundane time: 5 minutes, 15 minutes,
etc. However, it is a natural by-product of unbroken attention that it makes apparent
a new, subtler form of continuity. Between each breath, there is the time it takes
for the length of the in-breath and out-breath themselves. If we break the in-breath,
for example, into three pieces (beginning, middle, end) it is then natural to find
that, after one has sustained continuous awareness of these subsections of the
breath, that these three pieces can each be broken down further into smaller
slices. And so on. It is in
this way that Anapanasati teaches the meditator to meditate on \textbf{all of time}.

\todo{add anapana slice diagram}

Anapana eliminates space. Continuity is the source of constantly-narrowing attention
on space. Anapana is mastery over time.

\subsubsection{Vipassana}

Vipassana is harder. It stops ignoring space and actually adds it back into the
equation. You move your attention throughout your entire body (which is the only
meditation object you have direct access to) with the narrow focus you trained
yourself to achieve using Anapana. As your apparent physical
existence begins to break apart, you’ll open up your attention to the whole body, all
at once (or as much as possible at once) but even after 10 Vipassana courses, I rarely
ever do anything approximating that because I’m quite a beginner. 🙂

Vedana provides a sort of backdoor to all contact at the other six sense
doors. Direct observation of Vedana can only be of our own body. By observing the
entire body and ``dividing and dissecting'' it with our attention, we eventually
observe all Vedana and, therefore, all contact activity at the sense doors. This is
all the space there is to meditate on. We are meditating on \textbf{all of space}.

Vipassana maintains the mastery of time and continuity demanded in Anapana, adding
all of space. Vipassana is mastery over time + space.

\subsubsection{Continuity}

On Day 8 of the introductory 10-day Vipassana course, a new meditator is introduced
to the idea of complete continuity: meditating all the time. In the last days of the
10-day course, the meditator is asked to meditate in all four postures (sitting,
standing, walking, lying down) during all activities (walking, eating, using the
toilet, bathing, washing, sleeping, etc.).

Old student courses such as the 3-day course, Satipatthana Sutta course, and 20-day
course emphasize this point more strictly, beginning in the Anapana period. The
notion of continuity is expanded to Anapana and the idea of potentially meditating 24
hours a day is introduced. As the meditator begins meditating in all positions, and
particularly while lying down, sleep is naturally replaced with
meditation. Eventually, the conscious/unconscious barrier is broken even during
hours of dreaming and hours of deep sleep. One becomes conscious of bodily sensation,
either through the breath or throughout the entire body, 24 hours a day --- a true
continuity.

Thus, the teaching fulfills our initial three criteria: Vipassana, practiced
continuously, is a meditation on all of space, all of time, all the time.

Vipassana is mastery over Spacetime.
